---
title: "Math 430 Fall 2025"
format: 
  html:
    toc: true
---

## Announcement

- We will have a midterm redo **October 28**. 
- This is **optional**, you do not need to take it if you would like to keep your midterm grade.
- You **can** use one-sided standard sized (11 x 8.5in) piece of paper with notes on the midterm redo.
- You can make up to a 50% improvement on your midterm grade with the midterm redo.
  - This means if you got a 50% on the midterm originally, and a 100% on the redo midterm, your new midterm grade would be 50 + 0.5 * (100 - 50) = 75%
    
```{r}
original_midterm <- 50
redo_midterm <- 100

new_midterm_grade <- original_midterm + 0.5 * (redo_midterm - original_midterm)

new_midterm_grade
```

  - This means if you got a 50% on the midterm originally, and a 75% on the redo midterm, your new midterm grade would be 50 + 0.5 * (75 - 50) = 62.5%
    
```{r}
original_midterm <- 50
redo_midterm <- 75

new_midterm_grade <- original_midterm + 0.5 * (redo_midterm - original_midterm)

new_midterm_grade
```

  - This means if you got a 50% on the midterm originally, and a 30% on the redo midterm, your new midterm grade would be 50 + 0.5 * (30 - 50) = 40%
    
```{r}
original_midterm <- 50
redo_midterm <- 30

new_midterm_grade <- original_midterm + 0.5 * (redo_midterm - original_midterm)

new_midterm_grade
```
    


## HW 6

- [HW 6](https://csuci-math430.github.io/lectures/week-08/hw-06.qmd)
- [HW 6 data](https://csuci-math430.github.io/lectures/data/donner-party.csv)
- You can email me to submit your homework if Canvas is still down tomorrow.

## Syllabus

- [Syllabus](https://docs.google.com/document/d/1mgDjIgYG-hoV6fMIc6Yw9dvmGBToZOA-V5c0U1OUJx8/edit?usp=sharing)

## Lectures

- [Lecture 1a](https://csuci-math430.github.io/lectures/week-01/lecture-01a-introduction-to-class.html)
- [Lecture 1b](https://csuci-math430.github.io/lectures/week-01/lecture-01b-introduction-to-toolkit.html)
- [Lecture 2a](https://csuci-math430.github.io/lectures/week-02/lecture-02a-foundations-for-statistical-inference.html)
- [Lecture 2b](https://csuci-math430.github.io/lectures/week-02/lecture-02b-foundations-for-statistical-inference-part2.html)
- [Lecture 3a](https://csuci-math430.github.io/lectures/week-03/lec-03a-summary-stats-and-data-viz.html)
- Lecture 3b was practical 1
- [Lecture 4a](https://csuci-math430.github.io/lectures/week-04/lec-04a-intro-to-simple-linear-regression.html)
- [Lecture 4b](https://csuci-math430.github.io/lectures/week-04/lec-04b-simple-linear-regression-inference.html)
- [Lecture 5a](https://csuci-math430.github.io/lectures/week-05/lec-05a-linear-regression-conditions.html)
- [Lecture 5b](https://csuci-math430.github.io/lectures/week-05/lec-05b-model-evaluation.html)
- [Lecture 6a](https://csuci-math430.github.io/lectures/week-06/lec-06a-multiple-linear-regression.html)
- [Lecture 6b](https://csuci-math430.github.io/lectures/week-06/lec-06b-multiple-linear-regression-part2.html)
- Week 7 was the midterm
- [Lecture 8a](https://csuci-math430.github.io/lectures/week-08/lec-08a-logistic-regression.html) & [Annotated Lecture 8a](https://csuci-math430.github.io/lectures/week-08/lec-08a-logistic-regression-annotated.pdf)
- [Lecture 8b](https://csuci-math430.github.io/lectures/week-08/lec-08b-logistic-regression-part2.html)


