---
title: "Math 430: Lecture 8b"
subtitle: "Logistic regression part 2"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false

library(tidyverse)
library(janitor)
```

## Logisitic regression

$$\log\left(\frac{E[Y_i | X_{1i}, ..., X_{pi}]}{1 - E[Y_i | X_{1i}, ..., X_{pi}]}\right) = \beta_0 + \beta_1 X_{1i} + ... + \beta_p X_{pi}$$

We often let $\pi_i = E[Y_i | X_{1i}, ..., X_{pi}]$ so our model is written

$$\log\left(\frac{\pi_i}{1 - \pi_i}\right) = \beta_0 + \beta_1 X_{1i} + ... + \beta_p X_{pi}$$

# Interpreting coefficients

##

$\beta_0$ is the log odds for when all $X_i$'s are zero.

## 

$\beta_1$ is the log odds ratio associated with a one unit increase in $X_1$

- $e^{\beta_1}$ is the odds ratio associated with a one unit increase in $X_1$
- $e^{\beta_1}$ is the multiplicative change / factor change in odds associated with a one unit increase in $X_i$
- For $e^{\beta_1} > 1$, $100 * (e^{\beta_1} - 1)$% is the increase in odds associated with a one unit increase in $X_1$. For $e^{\beta_1} < 1$, we have $100 * (1 - e^{\beta_1})$% decrease in odds associated with a one unit increase in $X_1$

... with all other $X_i$'s held constant.

# Inference

In our class we will focus on two methods for hypothesis tests and confidence intervals

- Wald
- Likelihood ratio

## Comparing inference methods

::::{.columns}
:::{.column width="50%"}
**Wald** - For a single predictor*

- Pro: Easy to calculate
- Pro: Only requires fitting one model
- Con: Invalid for small sample size
- Con: Confidence intervals may contain invalid values outside [0,1]

:::

:::{.column width="50%"}
Likelihood ratio - similar to partial F / F-test for linear regression

:::
::::


## Comparing inference methods

::::{.columns}
:::{.column width="50%"}
**Wald** - For a single predictor

- Pro: Easy to calculate
- Pro: Only requires fitting one model
- Con: Invalid for small sample size
- Con: Confidence intervals may contain invalid values outside [0,1]

:::

:::{.column width="50%"}
**Likelihood ratio** - Compares nested models
  
- Pro: Confidence intervals will only contain valid values
- Pro: Not invalidated by small sample size
- Con: Requires fitting two models
- Con: "Not as easy to fit"

:::
::::

## Coronary heart disease (cdh) data

A retrospective sample of males in a heart-disease high-risk region of the Western Cape, South Africa. 
These data are taken from a larger dataset, described in  Rousseauw et al, 1983, South African Medical Journal.


## 

- `sbp`		systolic blood pressure
- `tobacco`		cumulative tobacco (kg)
- `ldl`		low density lipoprotein cholesterol
- `adiposity` body fat
- `famhist`		family history of heart disease
- `typea`		type-A behavior
- `obesity` 
- `alcohol`		current alcohol consumption
- `age`		age at onset

##

```{r}
heart_data <- read_csv(
  "https://hastie.su.domains/ElemStatLearn/datasets/SAheart.data", 
  show_col_types = FALSE
)

glimpse(heart_data)
```

## Fitting the model

\begin{equation*}
  \begin{split}
    \log(\frac{\pi_i}{1 - \pi_i}) &= \beta_0 + \beta_1 X_{\text{Blood pressure}, i}\\
    &+ \beta_2 X_{\text{Family history}, i} + \beta_1 X_{\text{Age}, i}
  \end{split}
\end{equation*}

```{r}
heart_model <- glm(
  chd ~ sbp + famhist + age,
  data = heart_data,
  family = binomial # This tells it to do logisitic regression
)
```

##

```{r}
summary(heart_model)
```

# Testing example 1

Let's test if systolic blood pressure should be in the model

$H_0: \beta_1 = 0$

$H_A: \beta_1 \neq 0$

We can do this with a Wald or Likelihood ratio test

```{r}
heart_full_model <- glm(
  chd ~ sbp + famhist + age,
  data = heart_data,
  family = binomial # This tells it to do logisitic regression
)
```


## Performing Wald Test

```{r}
summary(heart_full_model)
```

*The `summary()` function displays Wald Statistics.*

## Performing Likelihood Ratio Tests

```{r}
heart_reduced_model <- glm(
  chd ~ famhist + age,
  data = heart_data,
  family = binomial # This tells it to do logisitic regression
)

anova(heart_reduced_model, heart_full_model, test = "LRT")
```

## 

- The p-value for Wald test: 0.235
- The p-value for the Likelihood Ratio Test: 0.2344
- The two p-values won't necessarily equal the same. In practice you would choose one test to perform, you wouldn't fit both.

A probability of 0.02344 is not unlikely, so fail to reject $H_0$.

We found no evidence (Likelihood ratio test p-value = 0.2344) that the model with systolic blood pressure significantly improves the fit compared to the model with just family history and age.

# Testing example 2

Let's test if systolic blood pressure and family history should be in the model

$H_0: \beta_1 = \beta_2 = 0$

$H_A:$ At least one of $\beta_1$ or $\beta_2$ are zero

We can do this with a Wald or Likelihood ratio test


## Wald test

```{r}
library(aod)

wald.test(
  Sigma = vcov(heart_full_model), 
  b = coef(heart_full_model), 
  Terms = c(1, 2)
)
```

. . .

So we can test subsets of predictors with a Wald test, but it is different...

## Likelihood ratio test
```{r}
heart_reduced_model2 <- glm(
  chd ~ age,
  data = heart_data,
  family = binomial
)

anova(heart_reduced_model2, heart_full_model, test = "LRT")
```

##

Let's use the likelihood ratio test.

We found very strong evidence (likelihood ratio test p-value 3.9e-05) that including systolic blood pressure and family history improve the model fit, over a model with age only.


## Which to use?

. . .

- We already listed the pros and cons, so with those considerations you have to make a choice.
- I think it is important that you specify what test you used when providing the p-value.
- If you have a small sample size DON'T use Wald

## Confidence intervals 

Via Likelihood Ratio: We can used the `confint()` function like we did with linear regression

```{r}
confint(heart_full_model)
```

Via Wald: We can used the `confint.default()` function

```{r}
confint.default(heart_full_model)
```



# 

*We didn't have time to cover prediction in class so that was moved to Tuesday lecture 9a.*