---
title: "Math 430: Lecture 11a"
subtitle: "Overfitting and Cross Validation"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false

library(tidyverse)
library(janitor)

theme_set(theme_bw(base_size = 24))

penguins_cleaned <- penguins |> 
  drop_na()
```

## Overfitting

- We are fitting the model to sample data.

- Our goal is to understand the population data.

- If we make our model too perfect for our sample data, the model may not perform as well with other sample data from the population.

- In this case we would be overfitting the data.

- We can use **model validation** techniques.

## Data

We will be using the penguins_cleaned data to try to build a predictive model for penguin bill depth.

```{r}
ggplot(penguins_cleaned, aes(x = flipper_len, y = bill_dep, color = species)) +
  geom_point() +
  labs(x = "Flipper length (mm)", y = "Bill length (mm)", color = "Species")
```

# Method 1: Validation

To avoid overfitting we can split our data **validation**

1. Split sample of $n$ observations into two parts
2. Fit model with **training data** containing $m$ observations
3. Predict $Y$ for observations in **test data** containing $n - m$ observations, with model fit on training data
4. Evaluate prediction error on test data

## Splitting the data (Train vs. Test)

```{r message=FALSE}
set.seed(8236)
library(rsample)

penguins_cleaned_split <- initial_split(penguins_cleaned, prop = 0.75) ## 75% to 25% split
```

. . .

```{r}
dim(penguins_cleaned)
```

. . .

```{r}
penguins_cleaned_train_data <- training(penguins_cleaned_split)
dim(penguins_cleaned_train_data)
```

. . .

```{r}
penguins_cleaned_test_data <- testing(penguins_cleaned_split)
dim(penguins_cleaned_test_data)
```

##

Fit the model with the training data
```{r}
model_1_train <- lm(
  bill_dep ~ flipper_len + species + body_mass,  
  data = penguins_cleaned_train_data
)
```

Predict bill length for penguins_cleaned in testing data
```{r}
model_1_test_preditions <- predict(model_1_train, newdata = penguins_cleaned_test_data)
```

##

In sample (train) RMSE
```{r}
modelr::rmse(
  model_1_train, 
  data = penguins_cleaned_train_data
)
```

Our of sample (test) RMSE
```{r}
predicted_y <- model_1_test_preditions
actual_y <- penguins_cleaned_test_data$bill_dep

sqrt(mean((predicted_y - actual_y)^2, na.rm = TRUE))
```

Does it make sense that our in-sample RMSE > out-of-sample RMSE?

## Method 1: Validation - Summary

The validation approach 

Advantage:

- allows us to assess out-of-sample prediction error

Disadvantage:

- validation estimate of test error rate can be highly variable
- can result in worse model fit due to decreased sample size => overestimate test error rate


# Method 2: Leave-one-out Cross-Validation (LOOCV)

1. Split sample of $n$ observations into two parts
2. Fit model with **training data** containing $n-1$ observations
3. Predict $Y$ for observations in **test data** containing $1$ observation, with model fit on training data
3. Repeat steps 1-4 to get $CV_{(n)} = \frac{1}{n} \sum_{i = 1}^{n} (y_i - \hat{y}_i)^2$

##

Advantages:

- Training model on more data => less bias
- Consistent results since there is no randomness in data split

Disadvantages:

- Can be computation or time intensive with large data set

# Method 3: k-Fold Cross-Validation

1. Split sample of $n$ observations into $k$ groups of approximately equal size
2. First fold is test set and model is trained/fit on remaining $k-1$ sets
3. Predict $Y$ for observations in test data 
4. Evaluate prediction error on test data
5. Repeat $k$ times and calculate $CV_{(k)} = \frac{1}{k}\sum_{i=1}^{k} (y_i - \hat{y_i})^2$

##

Advantages:

- $k$-fold is less computationally intensive, relative to LOOCV, since the model only needs to be fit $k$ times, typically $k=5$ or $k=10$.
- Less variability than basic validation approach
- Often gives more accurate estimates of test error rate than LOOCV

##

```{r}
library(caret)
train_control <- trainControl(method = "cv", number = 10)

# Train a model (e.g., a linear model) using cross-validation
model <- train(
  bill_dep ~ flipper_len + species + body_mass,  
  data = penguins_cleaned_train_data, 
  method = "lm", 
  trControl = train_control
)

# View the results of the cross-validation
print(model)
```


## For classification

These methods can also be used for classification, but $(y_i - \hat{y}_i)^2$ would be replaces with $I(y_i \neq \hat{y}_i)$

 