---
title: "Math 430: Lecture 4b"
subtitle: "Simple Linear Regression for Inference"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
library(tidyverse)

theme_set(theme_bw(base_size = 24))
```


# Understanding the variation in our estimates

## Theoretical population regression line

We said our theoretical model is 
$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i.$$

It is theoretical because we are saying if we had all of the data for our population, a census instead of a sample, it would be the best linear approximation for the relationship between $X$ and $Y$.

Realistically we will almost never have a census, so $\beta_0$ and $\beta_1$ are unknown in practice.

## Simulation: population data

```{r}
#| echo: false

set.seed(3775)

pop_n <- 2500
sigma <- 2
beta_0 <- 2
beta_1 <- 3

sample_n <- 100

pop_data <- data.frame(
  x = runif(n = pop_n, min = -2, max = 2),
  epsilon = rnorm(n = pop_n, mean = 0, sd = sigma)
) |> 
  mutate(y = beta_0 + beta_1 * x + epsilon)
```

I simulated data for a population of `r pop_n` individuals.

1. randomly generate $x$'s (doesn't matter how).
2. randomly generate $\epsilon$'s where $\epsilon_i \overset{iid}{\sim} N(0, 1.5^2)$.
3. calculate $y$'s with $y_i = 3 + 2 x_i + \epsilon_i$.

```{r}
#| fig-align: center
#| echo: false
ggplot(pop_data, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## Simulation: simple random sample 1

```{r}
#| echo: false
#| eval: false

set.seed(2359)

sample_data <- NULL
estimate_df <- NULL

for (i in 1:1000) {
  sample_ids <- sample(pop_n, size = sample_n, replace = FALSE)

  sample_data <- pop_data |> 
    select(-epsilon) |> 
    slice(sample_ids) |> 
    mutate(sample_id = i) |> 
    bind_rows(sample_data)

  lm_model <- lm(y ~ x, data = sample_data)

  estimate_df <- data.frame(
    estimates = lm_model$coefficients,
    type = c("beta 0 hat", "beta 1 hat"),
    sample_id = i
  )  |> 
    bind_rows(estimate_df)
}

write_csv(sample_data, "../data/sim-sample-data.csv")
write_csv(estimate_df, "../data/sim-estimate-df.csv")
```

```{r}
#| echo: false

sample_data <- read_csv("../data/sim-sample-data.csv", show_col_types = FALSE)
estimate_df <- read_csv("../data/sim-estimate-df.csv", show_col_types = FALSE)
```

```{r}
#| echo: false

produce_sim_sample_plots <- function(sample_num) {
  gridExtra::grid.arrange(
    sample_data |> 
      filter(sample_id == sample_num) |> 
      ggplot(aes(x = x, y = y)) +
      geom_point() +
      geom_abline(intercept = beta_0, slope = beta_1, color = "blue", size = 1.5) +
      geom_smooth(method = "lm", se = FALSE, color = "red"),
    estimate_df |> 
      filter(sample_id <= sample_num) |> 
      ggplot(aes(x = estimates)) +
      geom_histogram(binwidth = 0.025, color = "black", fill = "grey") +
      geom_vline(
        data = data.frame(
          estimates = c(beta_0, beta_1),
          type = c("beta 0 hat", "beta 1 hat")
        ),
        mapping = aes(xintercept = estimates),
        color = "blue",
        size = 2
      ) +
      facet_wrap(vars(type), scales = "free_x") +
      ggh4x::facetted_pos_scales(x = list(
        type == "beta 0 hat" ~ scale_x_continuous(limits = c(
          filter(estimate_df, type == "beta 0 hat") |> pull(estimates) |> min(), 
          filter(estimate_df, type == "beta 0 hat") |> pull(estimates) |> max()
        )),
        type == "beta 1 hat" ~ scale_x_continuous(limits = c(
          filter(estimate_df, type == "beta 1 hat") |> pull(estimates) |> min(), 
          filter(estimate_df, type == "beta 1 hat") |> pull(estimates) |> max()
        ))
      )), 
    nrow = 2
  )
}
```

Next I randomly sampled `r sample_n` individuals and  calculate the least squares estimates $\hat{\beta}_0$ and $\hat{\beta}_1$

```{r}
#| fig-align: center
#| echo: false

produce_sim_sample_plots(1)
```

## Simulation: simple random sample 2

I can do that again with another random sample of `r sample_n`.

```{r}
#| fig-align: center
#| echo: false

produce_sim_sample_plots(2)
```

## Simulation: simple random sample 3

And again with a third random sample of `r sample_n`


```{r}
#| fig-align: center
#| echo: false

produce_sim_sample_plots(3)
```

## Simulation: simple random sample 100

And again 100 times!


```{r}
#| fig-align: center
#| echo: false

produce_sim_sample_plots(100)
```

## Simulation: simple random sample 1000

Even 1000 times!


```{r}
#| fig-align: center
#| echo: false

produce_sim_sample_plots(1000)
```
