---
title: "Math 430: Lecture 6b"
subtitle: "Multiple linear regression"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false

library(tidyverse)
theme_set(theme_bw(base_size = 24))
```


## Some important questions

When we perform multiple linear regression, we usually are interested in answering a few important questions.

1. Is at least one of the predictors $X_1, X_2,...,X_p$ useful in predicting the response?
2. Do all the predictors help to explain $Y$ , or is only a subset of the predictors useful?
3. How well does the model fit the data?
4. Given a set of predictor values, what response value should we predict, and how accurate is our prediction?

# Q1: Is There a Linear Relationship Between the Response and Predictors?

## Q1 with simple linear regression

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$$

. . .

Hypothesis test for $\beta_1$!

. . .

- $H_0: \beta_1 = 0$
- $H_A: \beta_1 \neq 0$
- Compute p-value. 

. . .

- *What is the definition of p-value?*

. . .

- Judge if you (1) found evidence against the null hypothesis (in support of the alternative) or (2) failed to find evidence against the null hypothesis (failed to find evidence to support the alternative)


## Q1 with multiple linear regression

$$Y_i = \beta_0 + \beta_1 X_{1i}  + \beta_2 X_{2i} + ... + \beta_p X_{p,i} + \epsilon_i$$

. . .

Hypothesis test for all $\beta_j$'s!

. . .

- $H_0: \beta_1 = \beta_2 = \dots = \beta_p = 0$
- $H_A$: At least one $\beta_j$ is non-zero
- Compute p-value. 
- Judge if you (1) found evidence against the null hypothesis (in support of the alternative) or (2) failed to find evidence against the null hypothesis (failed to find evidence to support the alternative)


## F-statistic

- $H_0: \beta_1 = \beta_2 = \dots = \beta_p = 0$
- $H_A$: At least one $\beta_j$ is non-zero

Our test test statistic for the hypothesis test is: 
$$F = \frac{(TSS - RSS) / p}{RSS / (n - p -1)}$$

. . .

- TSS = ...
- RSS = ...
- n = ...
- p = ...


## F-statistic (in practice)

The dataset contains data on test performance, school characteristics and student demographic backgrounds for school districts in California in 1999.

```{r}
ca_schools <- read_csv(
  "http://csuci-math430.github.io/lectures/data/CASchools.csv",
  show_col_types = FALSE
)

ca_schools <- ca_schools |> 
  mutate(student_teacher_ratio = students / teachers) |> 
  mutate(student_computer_ratio = case_when(
    computer != 0 ~ students / computer,
    computer == 0 ~ 0
  )) |> 
  mutate(ave_score = (read + math) / 2)
```


## `CASchools` data

::::{.columns}

:::{.column width="50%"}
```{r}
#| echo: false
ggplot(
  ca_schools, 
  aes(x = ave_score)
) +
  geom_histogram() +
  labs(
    x = "Average test score",
    y = "Count"
  )
```

:::

:::{.column width="50%"}

```{r}
#| echo: false
ggplot(
  ca_schools, 
  aes(x = student_teacher_ratio, y = ave_score)
) +
  geom_point() +
  labs(
    x = "Student-to-Teacher Ratio",
    y = "Average test score"
  )
```
:::

::::

::::{.columns}

:::{.column width="50%"}

```{r}
#| echo: false
ggplot(
  ca_schools, 
  aes(x = student_computer_ratio, y = ave_score)
) +
  geom_point() +
  labs(
    x = "Student-to-Computer Ratio",
    y = "Average test score"
  )
```
:::

:::{.column width="50%"}

```{r}
#| echo: false
ggplot(
  ca_schools, 
  aes(x = lunch, y = ave_score)
) +
  geom_point() +
  labs(
    x = "% Qualifying for reduced-price lunch",
    y = "Average test score"
  )
```
:::
::::

