---
title: "Math 430: Lecture 15b"
subtitle: "Handling Missing Data"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
options(scipen = 0)
library(tidyverse)
library(openintro)
library(broom)
library(modelr)
theme_set(theme_bw(base_size = 22))
```

## Missing data

In R, missing data are indicated by `NA`'s. 

When plotting or fitting models, any row with missing data will not be used, reducing the sample size $n$. 

- This is called **complete-case analysis / listwise deletion**

## Concerns for dropping data

When doing prediction

- Without complete-case analysis: If we compare models with different predictors, we may be comparing different sample size $n$ as well
- With complete-case analysis: We are losing some information that may have improved our prediction

. . .

When doing inference (prespecified single model)

- We want as large a sample size as possible to be able to detect differences/associations (lower standard errors)

# Missing-data mechansims

*What if the reason some observations are missing is related to the phenomena we are trying to model?*

## Missing Completely at Random (MCAR) 

MCAR is when the probability of missingness is the same for all units of observation (e.g., people in a clinical trial have the same probability of having missingness for any variable measured)

- complete-case analysis does not bias results in this case


## Missing at random (MAR)

MAR is when the missingness is not completely at random, but instead due to the variables we observed.

- complete-case analysis can still be done here, as long as the model contains all variables that explain the missing data

## Missing not at random (MNAR)

MNAR is when the missingness is not MCAR or MAR

- Today there are even finer-grained distinctions of different reasons for MNAR
- Ex: A familiar example from medical studies is that if a particular treatment causes
discomfort, a patient is more likely to drop out of the study
- Not accounting for the  missingness, like doing complete-case analysis will *result in biased inference!*

## 

We will not focus on MNAR in this discussion, instead let's discuss how we can still use all the data we have for MCAR and some MAR scenarios

. . .

- If we could predict the missing values, then proceed with our modeling
- => We would have a larger $n$ than complete-case analysis,
- => have smaller standard errors,
- => and more statistical power to detect differences/associations

# Imputation

**Imputation** is the process of using available information to create substitute values for missing values in our data, so we can use that data for analysis

- There are many imputation methods

## Mean imputation

- Idea: Replace missing values for a variable with the average observed value for that variable

- Ex: If some people did not report their age in a survey, we can replace `NA`'s with the average of the observed ages

- Pros: Dimple

- Cons: Can distort relationships and variances. Only works for numeric variables

## Example

We are very familiar with the penguins data
```{r}
glimpse(penguins)
```

. . .

```{r}
# n for complete-case analysis
penguins |> 
  drop_na() |> 
  nrow()
```

##

\begin{equation*}
  \begin{split}
    Bill Depth_i &= \beta_0 + \beta_1 Flipper Length_i\\
    &+ \beta_2 I(Species_i = Chinstrap)\\
    &+ \beta_3 I(Species_i = Gentoo) + \epsilon_i
  \end{split}
\end{equation*}

```{r}
sapply(penguins, function(x) sum(is.na(x)))
```

## Without imputation

```{r}
penguin_model1 <- lm(
  bill_dep ~ flipper_len + species, 
  data = penguins
)
summary(penguin_model1)
```

## Mean imputation of `bill_dep` and `flipper_len`

```{r}
penguins_model_data <- penguins |> 
  mutate(new_flipper_len = ifelse(
    is.na(flipper_len), 
    yes = mean(flipper_len, na.rm = TRUE),
    no = flipper_len
  )) |> 
  mutate(new_bill_dep = ifelse(
    is.na(bill_dep), 
    yes = mean(bill_dep, na.rm = TRUE),
    no = bill_dep
  ))
```

## Fitting model to imputated data

```{r}
penguin_model2 <- lm(
  new_bill_dep ~ new_flipper_len + species, 
  data = penguins_model_data
)
summary(penguin_model2)
```

##

```{r}
new_penguins <- penguins %>%
  mutate(across(
    where(is.numeric), 
    ~ ifelse(
      is.na(.), 
      mean(., na.rm = TRUE), 
      .
    )
  ))

glimpse(new_penguins)
```

