---
title: "Lab 1a: Familiarizing yourself with a Quarto document"
author: "Dr. Medina"
format: html
---

```{r}
library(tidyverse)
```

In this lab we will be exploring California unemployment rates.
This data is available on the [CA Open Data Portal](https://data.ca.gov/dataset/unemployment-rate-by-age-groups).

We will begin by reading in this data and saving it into an **object** called `unemployment`.

```{r}
unemployment <- read_csv("http://csuci-math430.github.io/lectures/week-01/data/unemployment_rate_by_age_groups.csv")
```

We can view what is in the `employment` object by printing it.

```{r}
unemployment
```

You can skip over the following code chunk, I needed to do some data cleaning and wrangling to get it in a good format for plotting.

```{r}
unemployment_new <- unemployment |>
  mutate(new_date = as_date(Date, format = "%m/%d/%Y")) |> 
  pivot_longer(
    cols = starts_with("Age "),
    names_to = "age_group",
    names_prefix = "Age ",
    values_to = "unemployment_rate"
  )
```


## Unemployment over time

```{r}
unemployment_new |> 
  filter(age_group == "25-34") |> 
  ggplot(aes(x = new_date, y = unemployment_rate)) +
  geom_point() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = "Date",
    y = "Unemployment rate \n for individuals ages 25-34"
  )
```

If you were to describe these trends to someone what would you say?

*You can type your response here*

Do you notice anything interesting?


Given this data, what would you predict the average unemployment rate of 2026 to be for people aged 25-34? Why?


Given this data, what would you predict the average unemployment rate of 2026 to be for people aged 55-64? Why?


## Unemployment trends for each age group

```{r}
ggplot(
  unemployment_new, 
  aes(x = new_date, y = unemployment_rate, color = age_group, linetype = age_group)
) +
  geom_line() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    x = "Date",
    y = "Unemployment rate",
    color = "Age group",
    linetype = "Age group"
  )
```

Which age group(s) had the highest unemployment rate in general?


Which age group(s) had the lowest unemployment rate in general?


Would you say there is a meaningful difference in unemployment rates between 55-64 year olds and 20-24 year olds? Why?

Would you say there is a meaningful difference in unemployment rates between 55-64 year olds and 35-34 year olds? Why?
