---
title: "Math 430: Lecture 13b"
subtitle: "Selecting variables for prediction modeling"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
options(scipen = 0)
library(tidyverse)
library(openintro)
library(broom)
library(modelr)
theme_set(theme_bw(base_size = 22))
```

## What variables to include...

We don't want to include extra variables as features in our predictive regression models if they do not improve the predictive performance. 

- This is where **subset selection** comes in, to help us choose our features ($X$'s).

. . .

On the other hand, we may have very large number of predictors $p$ available, relative to the sample size $n$. 

- In very high dimensional data situations, **dimension reduction** can help by tranforming and combining features

## Subset selection

- Best subset selection
- Stepwise forwards  selection
- Stepwise backwards selection

##

```{r}
data("swiss")

glimpse(swiss)
```

## Standardizing our variables

When we are doing subset selection we should standardize our variables (subtract mean & divide by standard deviation) to avoid comparing variables of different magnitudes.

The `scale()` function standardizes columns, and we can tell it to only attempt for numeric columns:
```{r}
swiss_prepped <- swiss %>%
  mutate(across(where(is.numeric), scale))
```


## Best subset selection idea

Fit a model for each possible combination and compare models using appropriate metric

- Need to use a metric appropriate for comparing models of varying parameters (i.e. NOT $R^2$)

## Best subset selection steps

1. Fit *null model*, model with no predictors.
2. For each $k = 1, 2, ..., p$:
  a. Fit all ${p}\choose{k}$ models with $k$ predictors
  b. Pick the best for that $k$
3. Select *best* model among the $p+1$ options using out-of-sample prediction error RMSE / Accuracy or $C_p$.


## Best subset selection considerations

Pros:

- Finds optimal combination of predictors

Cons:

- Must fit many models, $2^p$, can be computationally expensive for large $p$ or $n$

## Forward stepwise selection idea

Begins with a model with no predictors, then adds predictors one at a time, keeping the one that gives the greatest additional improvement each round. Stops when no variable improves model

## Forward stepwise selection steps

1. Fit *null model*, model with no predictors.
2. For each $k = 0, 1, ..., p-1$:
  a. Add one additional $p-k$ predictor to current "best" model
  b. Choose best among these models
3. Select *best* model among the $p+1$ options using out-of-sample prediction error RMSE / Accuracy or $C_p$.

## Forward stepwise selection considerations

Pros:

- Fits total of $1 + \sum_{k=0}^{p-1} (p - k) = 1 + p(p+1)/2$ models => computational advantage over best subset selection
- Works even when $p>n$

Cons:

- Not guaranteed to yield the best model containing a subset of $p$ predictors
  - May get stuck at local optima


## Backwards stepwise selection idea

Opposite of forwards selection, begins with full model using all available predictors, then iteratively removes one each step.

## Backwards stepwise selection steps

1. Fit full model, model with all predictors.
2. For each $k = p, p-1, ..., 1$:
  a. Consider all $k$ models that contain all but one of the predictors in the current baseline model, for a total of $k-1$ predictors
  b. Choose best among these.
3. Select *best* model among the $p+1$ options using out-of-sample prediction error RMSE / Accuracy (or $C_p$...).


## Backwards stepwise selection considerations

Pros:

- Also fits total of $1 + p(p+1)/2$ models => computational advantage over best subset selection

Cons:

- Not guaranteed to yield the best model containing a subset of $p$ predictors
  - May be better than forwards for exploring
- Doesn't work when $p>n$

## Overall

- Best subset, forward stepwise, and backwards stepwise give fairly similar but not identical models typically
- Sequential stepwise selection is a combination of forward and backward, where a variable can be removed or added at each stage
- Step 3 should be rigorously assessed using validation approaches, but some metrics can be used in cases of low sample size

## $C_p$

What if we don't want to split our data to do validation? $C_p$ provides an sense of test-MSE for linear regression

$$C_p = \frac{1}{n} (RSS + 2d \hat{\sigma}^2),$$

where $\hat{\sigma}^2$ is an estimate of the variance of the error $\epsilon$, typically estimated using full model. 

There are other similar metrics like AIC, BIC, etc.

**With today's computational power and support we should prefer validation methods!**

## Doing best subset selection with $C_p$

Note `method` can be set to "forwards", "backwards", or "sequential"

```{r}
library(leaps) # for regsubsets() function

model_best <- regsubsets(
  Fertility ~ ., # Period means to use all of the variables
  data = swiss_prepped, 
  method = "exhaustive" # For best subset selction
)

summary(model_best)
```

## Extracting evaluation metrics

```{r}
model_best_summary <- summary(model_best)
model_best_summary$cp
```

Using $C_p$ instead of k-fold cross validation we see it selects for the fourth model.

## Doing backwards/forwards selection with k-fold cross validation

The caret package offer help with 

- forwards feature selection via k-fold cross validation (Selection by Filter) with `sbf()`
- backwards feature selection via k-fold cross validation (Recursive feature elemination) with `rfe()`

```{r}
library(caret)
```

## Backwards

```{r}
rfe_control <- rfeControl(
  functions = lmFuncs, # lmFuncs, glmFuncs, etc., depending on your model
  method = "cv",       # k-fold cross validation
  number = 10         # Number of folds
)

rfe_result <- rfe(
  Fertility ~ ., 
  data = swiss_prepped,
  metric = "RMSE",
  rfeControl = rfe_control
)

predictors(rfe_result)
```

## Forwards

```{r}
sbf_control <- sbfControl(
  functions = lmSBF,
  method = "cv",       # k-fold cross validation
  number = 10         # Number of folds
)

sbf_result <- sbf(
  Fertility ~ ., 
  data = swiss_prepped,
  metric = "RMSE",
  sbfControl = sbf_control
)

predictors(sbf_result)
```

## Takeaways

- Difference between subset selection and dimensionality reduction
- Forward/backward stepwise selection is typically more practical than best subset selection (just choose one)
- We should use k-fold cross validation for subset selection

