---
title: "Math 430: Lecture 13a"
subtitle: "Subset selection"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
options(scipen = 0)
library(tidyverse)
library(openintro)
library(broom)
library(modelr)
theme_set(theme_bw(base_size = 22))
```

# Variable selection for inference

*In class we took a while reviewing lecture 12b.*

*We reviewed variable types and including interactions.*

## Reviewing interactions

Regardless of study type (observational/experiment), if we miss a mediating variable we may think the relationship is completely different from the truth.

## Without species included

There appears to be a negative relationship between flipper length and bill depth...

```{r}
ggplot(penguins, aes(x = flipper_len, y = bill_dep)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

## With species included

In reality there is a positive relationship between flipper length and bill depth that differs by species.

```{r}
ggplot(penguins, aes(x = flipper_len, y = bill_dep, color = species)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

##

We can account for varying y-intercepts by adding species to our model, we can account for varying slopes by including interactions between species and flipper length.
\begin{equation*}
  \begin{split}
    &Bill Depth_i = \beta_0\\
    &+ \beta_1 Flipper Length\\
    &+ \beta_2 I(species_i = Chinstrap)\\
    &+ \beta_3 I(species_i = Gentoo)\\
    &+ \beta_4 I(species_i = Chinstrap) \times Flipper Length\\
    &+ \beta_5 I(species_i = Gentoo) \times Flipper Length + \epsilon_i
  \end{split}
\end{equation*}


##

```{r}
model <- lm(bill_dep ~ flipper_len * species, data = penguins)
summary(model)
```


## Fitted model

```{r}
#| echo: false

betas <- round(model$coefficients, digits = 2)
```


\begin{equation*}
  \begin{split}
    &\widehat{Bill Depth}_i = `r betas["(Intercept)"]` \\
    &+ `r betas["flipper_len"]` Flipper Length\\
    &+ `r betas["speciesChinstrap"]` I(species_i = Chinstrap)\\
    &+ `r betas["speciesGentoo"]` I(species_i = Gentoo)\\
    &+ `r betas["flipper_len:speciesChinstrap"]` I(species_i = Chinstrap) \times Flipper Length\\
    &+ `r betas["flipper_len:speciesGentoo"]` I(species_i = Gentoo) \times Flipper Length
  \end{split}
\end{equation*}

We fit a single model to leverage the larger sample size, instead of fitting an individual model per species.

# Subset selection

When doing prediction with regression we 

- want to include available data as predictors/feature, when it will help improve our prediction.
- are not concerned about multiple testing, because we aren't doing inference.
- don't want to check every possible combination manually.

## Subset selection

- Best subset selection
- Stepwise forwards  selection
- Stepwise backwards selection

. . .

*For all of these, reserving some data for testing is a good idea.*

## Best subset selection idea

Fit a model for each possible combination and compare models using appropriate metric (e.g., RMSE, $C_p$, adjusted $R^2$)

## Best subset selection steps

1. Fit *null model*, model with no predictors.
2. For each $k = 1, 2, ..., p$:
  a. Fit all ${p}\choose{k}$ models with $k$ predictors
  b. Pick the best for that $k$, using RSS / deviance
3. Select *best* model among the $p+1$ options using out-of-sample prediction error RMSE / Accuracy or $C_p$.

## Best subset selection considerations

- Must fit many models, $2^p$, can be computationally expensive for large $p$ or $n$
- Can overfit to training data if not splitting data




