---
title: "Math 430: Lecture 13a"
subtitle: "Subset selection"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
options(scipen = 0)
library(tidyverse)
library(openintro)
library(broom)
library(modelr)
theme_set(theme_bw(base_size = 22))
```

##

When doing prediction with regression we 

- want to include available data as predictors/feature, when it will help improve our prediction.
- are not concerned about multiple testing, because we aren't doing inference.
- don't want to check every possible combination manually.

## Subset selection

- Best subset selection
- Stepwise forwards  selection
- Stepwise backwards selection

. . .

*For all of these, reserving some data for testing is a good idea.*

## Best subset selection idea

Fit a model for each possible combination and compare models using appropriate metric (e.g., RMSE, $C_p$, adjusted $R^2$)

## Best subset selection steps

1. Fit *null model*, model with no predictors.
2. For each $k = 1, 2, ..., p$:
  a. Fit all ${p}\choose{k}$ models with $k$ predictors
  b. Pick the best for that $k$, using RSS / deviance
3. Select *best* model among the $p+1$ options using out-of-sample prediction error RMSE / Accuracy or $C_p$.

## Best subset selection considerations

- Must fit many models, $2^p$, can be computationally expensive for large $p$ or $n$
- Can overfit to training data if not splitting data

## Forward stepwise selection idea

Begins with a model with no predictors, then adds predictors one at a time, keeping the one that gives the greatest additional improvement each round. Stops when no variable improves model

## Forward stepwise selection steps

1. Fit *null model*, model with no predictors.
2. For each $k = 0, 1, ..., p-1$:
  a. Add one additional $p-k$ predictor to current "best" model
  b. Choose best among these models using RSS / deviance, proceeding with that model as baseline for next round of $k$.
3. Select *best* model among the $p+1$ options using out-of-sample prediction error RMSE / Accuracy or $C_p$.

## Forward stepwise selection considerations

- Fits total of $1 + \sum_{k=0}^{p-1} (p - k) = 1 + p(p+1)/2$ models => computational advantage over best subset selection
- Not gauranteed to yield the best model containing a subset of $p$ predictors
- Works even when $p>n$

## Backwards stepwise selection steps

Opposite of forwards selection, begins with full model using all available predictors, then iteratively removes one each step.

##

1. Fit full model, model with all predictors.
2. For each $k = p, p-1, ..., 1$:
  a. Consider all $k$ models that contain all but one of the predictors in the current baseline model, for a total of $k-1$ predictors
  b. Choose best among these models using RSS / deviance, proceeding with that model as baseline for next round of $k$.
3. Select *best* model among the $p+1$ options using out-of-sample prediction error RMSE / Accuracy or $C_p$.

##

- Also fits $1 + p(p+1)/2$ models, like forwards selection
- Also not guaranteed to yield the best model containing a subset of $p$ predictors
- Does not work when $p > n$

## Overall

- Best subset, forward stepwise, and backwards stepwise give fairly similar but not identical models typically
- Step 3 should be rigorously assessed using validation approaches, but some metrics can be used in cases of low sample size

## $C_p$

For linear regression

$$C_p = \frac{1}{n} (RSS + 2d \hat{\sigma}^2),$$

where $\hat{\sigma}^2$ is an estimate of the variance of the error $\epsilon$, typically estimated using full model.

Adds penalty for number of predictors

Is large or small $C_p$ good?

##

```{r}
data("swiss")

glimpse(swiss)
```

## Doing best subset

```{r}
library(leaps)

models <- regsubsets(Fertility ~ ., data = swiss, method = "exhaustive")
summary(models)
```

## Doing forwards stepwise

```{r}
library(leaps)

models <- regsubsets(Fertility ~ ., data = swiss, method = "forward")
summary(models)
```

## Doing backwards stepwise

```{r}
library(leaps)

models <- regsubsets(Fertility ~ ., data = swiss, method = "backward")
summary(models)
```


