---
title: "Math 430: Lecture 5b"
subtitle: "Linear Regression Model Evaluations"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false

library(tidyverse)

elephants <- read_csv(
  "http://csuci-math430.github.io/lectures/data/elephants.csv",
  show_col_types = FALSE
)
```


## How do we decide one model is better than another?

. . .

It depends on our purpose.


There are three values commonly used:

- Residual Standard Error (RSE)
- Root Mean Squared Deviation (RMSE)
- $R^2$


## Residual Standard Error (RSE)

$$RSE = \hat{\sigma} = \sqrt{\frac{1}{n - 2} RSS} = \sqrt{\frac{1}{n - 2} \sum_{i = 1}^{n} (y_i - \hat{y}_i)^2}$$

- Measures *lack of fit* of the model to the data
- Measured in units of $y$

. . .

$n - 2$ is required to make $E[\hat{\sigma}] = \sigma$, but if often replaced by just $n$ when the sole goal is prediction accuracy instead of inference

- using $n$ instead of $n - 2$ is called **Root Mean Squared Error** (RMSE)


##

What would you say is a "good" RSE/RMSE value? What would be a "bad" one?

. . .

- You can evaluate it relative to RSE/RMSE of another model (a model with different $X$ for the same $Y$)
- RMSE is "roughly" the average error, so you can think in context if the average error is acceptable or unacceptable

## $R^2$ statistic

$$R^2 = \frac{TSS - RSS}{TSS}$$

where $TSS = \sum_{i = 1}^{n} (y_i - \bar{y}_i)^2$.

- $TSS$ measures amount of variability in $Y$
- $RSS$ measures amount of variability in $Y$ unexplained by model
- $R^2$ measures percent of variability in $Y$ explained by model
- $TSS - RSS$ measures amount of variability explained by the model

## $R^2$ statistic

*What* $R^2$ *be for perfectly linear data?*

*What would* $R^2$ *be for completely unlinear data?*

. . .

In practice it is context specific for what constitutes a "reasonably good" $R^2$

- consider physics versus marketing

## Correlation and the $R^2$ statistic

$Cor(X, Y) = \frac{\sum_{i = 1}^{n} (x_i - \bar{x})(y_i - \bar{y})}{\sqrt{\sum_{i = 1}^{n} (x_i - \bar{x})^2 }\sqrt{\sum_{i = 1}^{n} (y_i - \bar{y})^2}}$

Correlation between $X$ and $Y$ is a measure of the linear relationship between $X$ and $Y$

. . .

For simple linear regression $R^2 = [Cor(X, Y)]^2$

## Cautions with $R^2$statistic

Low $R^2$ does not mean no/(or weak) relationship, simply not a linear relationship

::::{.columns}
:::{.column width="80%"}
```{r}
#| echo: false

set.seed(2343)
bad_linear_data <- data.frame(
  x = runif(100, -10, 10),
  residual = rnorm(100, mean = 0, sd = 15)
) |> 
  mutate(y = 2*x + x^2 + 1 + residual)

r_squared <- cor(bad_linear_data$x, bad_linear_data$y)^2
r_squared <- round(r_squared, 2)

ggplot(bad_linear_data, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

:::

:::{.column width="20%"}
$R^2$ = `r r_squared`
:::
::::

## Cautions with $R^2$statistic

High $R^2$ does not mean a regression line fits the data well. Another function may describe it better

::::{.columns}
:::{.column width="80%"}
```{r}
#| echo: false
bad_linear_data <- data.frame(
  x = runif(100, -10, 10),
  residual = rnorm(100, mean = 0, sd = 300)
) |> 
  mutate(y = 2*x + 4 * x^3 + 1 + residual)

r_squared <- cor(bad_linear_data$x, bad_linear_data$y)^2
r_squared <- round(r_squared, 2)

ggplot(bad_linear_data, aes(x = x, y = y)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

:::

:::{.column width="20%"}
$R^2$ = `r r_squared`
:::
::::


# Estimating the response

## 

We know $\hat{y}_i$ is our model's point estimate for expectation of $Y_i$

What about an uncertainty estimate?

## Confidence interval for average $Y_i$

Let $\mu_Y = E[Y_i | X_i = x_h]$, the mean response when the explanatory variable value is $x_h$.

$$\text{Point estimate} \pm t^* \times SE(\text{Point estiamte})$$

. . .

$$\hat{y}_h \pm t_{(1 - \alpha / 2, n - 2)} RSE\sqrt{ \left(\frac{1}{n} + \frac{(x_h - \bar{x})^2}{\sum_{i = 1}^{n} (x_i - \bar{x})^2}\right)}$$

. . .

- $x_h$ must be within the range of the observed $x_i$ values to avoid extrapolation, but need not be an observed $x_i$ 


## Prediction interval for a new response

Let $\hat{y}_{new}$ be the new response when the explanatory variable value is $x_h$.

$$\hat{y}_h \pm t_{(1 - \alpha / 2, n - 2)} RSE \sqrt{1 + \frac{1}{n} + \frac{(x_h - \bar{x})^2}{\sum_{i = 1}^{n} (x_i - \bar{x})^2}}$$

- $x_h$ must be within the range of the observed $x_i$ values to avoid extrapolation, but need not be an observed $x_i$

. . .

- *How does this compare the the CI for average* $Y_i$*?*
 
## Example: Hospital infection risk

The hospital infection risk dataset consists of a sample of n = 58 hospitals in the east and north-central U.S. ([Hospital Infection Data Region 1 and 2 data](https://online.stat.psu.edu/onlinecourses/sites/stat501/files/data/hospitalinfct_reg1and2.txt)). 

- The response variable is y = infection risk (percent of patients who get an infection).
- The explanatory variable is x = average length of stay (in days).

## Example: Hospital infection risk data

```{r}
hospital_infection <- read_tsv("https://online.stat.psu.edu/onlinecourses/sites/stat501/files/data/hospitalinfct_reg1and2.txt")

glimpse(hospital_infection)
```

## Example: Hospital infection risk model

Fitting the model
```{r}
hospital_model <- lm(InfctRsk ~ Stay, data = hospital_infection)
summary(hospital_model)
```

## Example: CI for Average risk of infection for average stay of 10 days?

```{r}
new_hospital_data <- data.frame(Stay = 10)

predict(
  hospital_model, 
  newdata = new_hospital_data, 
  interval = "confidence",
  level = 0.95
)
```

```{r}
#| echo: false

y_ci <- predict(
  hospital_model, 
  newdata = new_hospital_data, 
  interval = "confidence",
  level = 0.95
)
```


We estimate the average risk of infection associated with a hospital with an average stay of 10 days to be `r round(y_ci[1], digits = 2)`. We estimate it to be between `r round(y_ci[2], digits = 2)` and `r round(y_ci[3], digits = 2)` with 95% confidence.

## Example: Prediction interval for risk of infection for average stay of 10 days?

```{r}
new_hospital_data <- data.frame(Stay = 10)

predict(
  hospital_model, 
  newdata = new_hospital_data, 
  interval = "prediction",
  level = 0.95
)
```

```{r}
#| echo: false

y_pi <- predict(
  hospital_model, 
  newdata = new_hospital_data, 
  interval = "prediction",
  level = 0.95
)
```


We estimate the risk of infection associated with a new hospital with an average stay of 10 days to be `r round(y_pi[1], digits = 2)`. We estimate it to be between `r round(y_pi[2], digits = 2)` and `r round(y_pi[3], digits = 2)` with 95% confidence.

## Confidence bands 

We can easily plot confidence bounds for average response - confidence intervals for average response for many $x_i$'s

```{r}
ggplot(hospital_infection, aes(x = Stay, y = InfctRsk)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE)
```

## Confidence bands 

```{r}
#| echo: false
#| fig-align: center
ggplot(hospital_infection, aes(x = Stay, y = InfctRsk)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  geom_segment(
    aes(
      x = 10, 
      xend = 10, 
      y = y_ci[2],
      yend = y_ci[3]
    ),
    color = "red",
    linewidth = 1.25,
    linetype = "dotted",
    alpha = 0.75
  )
```

The red line is our 95% confidence interval

## Confidence bands 

```{r}
#| echo: false
#| fig-align: center
ggplot(hospital_infection, aes(x = Stay, y = InfctRsk)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  geom_segment(
    aes(
      x = 10, 
      xend = 10, 
      y = y_ci[2],
      yend = y_ci[3]
    ),
    color = "red",
    linewidth = 1.25,
    linetype = "dotted",
    alpha = 0.75
  ) +
  geom_segment(
    aes(
      x = 10, 
      xend = 10, 
      y = y_pi[2],
      yend = y_pi[3]
    ),
    color = "purple",
    linewidth = 1.25,
    linetype = "dashed",
    alpha = 0.75
  )
```

The purple line is the 95% prediction interval for a new hospital