---
title: "Math 430: Lecture 5a"
subtitle: "Linear Regression Conditions"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false

set.seed(23243)
library(tidyverse)

elephants <- read_csv(
  "http://csuci-math430.github.io/lectures/data/elephants.csv",
  show_col_types = FALSE
)
```


##

We discussed how simple linear regression can be used to solve different goals

**Prediction** 

- We want the fitted line so we can use it on future data

**Inference**

- Learn the relationship between $Y$ and $X$ (confidence interval)
- Test if there is a linear relationship between $Y$ and $X$ (hypothesis test)


## 

Recall our linear regression from last week
```{r}
ggplot(elephants, aes(x = Height, y = Mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Height (cm)", y = "Mass (kg)")
```

. . .

*Was this model appropriate?*

## Why would we care if the model is "appropriate"?

::::{.columns}
:::{.column width="50%"}
```{r}
#| echo: false
ggplot(elephants, aes(x = Height, y = Mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Height (cm)", y = "Mass (kg)")
```

:::

:::{.column width="50%"}

- it may be chance that we got a good fit for our sample, but the model might be a bad fit for the population
- our estimates relied on our conditions, so if those don't hold our estimates are meaningless

:::
::::

## Conditions

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i, \epsilon_i \overset{iid}{\sim} N(0, \sigma^2)$$

In order of importance:

1. Linearity: Linear relationship between $X$ and $Y$ ($E[\epsilon_i] = 0)$.
3. Independence: Independence of $\epsilon_i$.
2. Homoskedasticity: Constant variance of $\epsilon_i$.
4. Normality: Normally distributed $\epsilon_i$.

# Checking model conditions

## Visual checks versus statistical tests

People have developed some statistical tests to asses some model conditions.

- We want to avoid unnecessary tests because each test we perform introduces some error.

. . .

Alternatively, we will use visual assessments

- There is some level of subjectivity
- We will use experience and reproducibility to be responsible

## Visual checks

```{r}
#| fig-align: center
#| output-location: slide
model_elephant <- lm(Mass ~ Height, data = elephants)

library(ggfortify)

autoplot(model_elephant)
```

Recall residuals: $e_i = y_i - \hat{y}_i$

## Diagnosing linearity

- Examine plot of residuals vs fitted (or predictor for simple linear regression)
- There should be no discernible pattern the the residuals centered around 0

:::{.pull-left}

**Linear**
```{r  echo = FALSE, message = FALSE}
set.seed(84735)
x <- seq(-2, 2, by = 0.01)
y <- 4*x + 5 + rnorm(length(x), 0 , 1.5)

data_good <- data.frame(x = x, y = y) |>   sample_n(50)

data_good |> 
  ggplot() +
  aes(x = x, y = y) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE)
```

:::

:::{.pull-right}

**Non-linear**

```{r  echo = FALSE, message = FALSE}
set.seed(84735)
x <- seq(-2, 2, by = 0.01)
y <- 3*x^2 + x + 5 + rnorm(length(x), 0 , 2)

data_bad <- data.frame(x = x, y = y) |> sample_n(50)

data_bad |> 
  ggplot() +
  aes(x = x, y = y) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  stat_function(fun = function(x)3*x^2 + x + 5 )
```

:::

## When linearity is violated

Implications: 

- poor fit
- biased inference

Remedy:

- consider transformations of $X$ or other relationships

## Diagnosing Independence

- Consider how the data were collected (random sample? no repeat sampling?) and on whom the data were collected (would you expect independence?)
- Dependence could be revealed by plotting residuals versus unmodeled variables (time, class, etc)

## When independence is violated

Implications:

- Estimators $\hat{\beta}_0$ and $\hat{\beta}_1$ are still unbiased
- The standard errors will underestimate the true standard errors
- Increased type one error rates (false positive)

Remedy:

- Consider a different model that accounts for correlation (e.g. if you have longitudinal data then a linear mixed effects model may be appropriate)

## Diagnosing homoskedasticity

- Examine plot of residuals vs fitted and check there is not a "funnel-shape"
- Specifically, for each fitted value does the spread of the residuals look similar?

:::{.pull-left}

**Constant variance**

```{r echo = FALSE, message = FALSE}
model_good <- lm(y ~ x, data = data_good)

data_good <- 
  data_good |>
  sample_n(50) |> 
  modelr::add_residuals(model_good) 

data_good |> 
  ggplot(aes(x = x, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0)

```

:::

:::{.pull-right}

**Non-constant variance**


```{r echo = FALSE, message = FALSE}
bad_data <- data.frame(
  x = seq(0, 10, length.out = 100)
) |> 
  mutate(y_perfect = 2 * x + 5) |> 
  mutate(residuals = rnorm(100, mean = 0, sd = abs(x) / 2)) |> 
  mutate(y = y_perfect + residuals)

model_bad <- lm(y ~ x, data = bad_data)

bad_data <- 
  bad_data |> 
  modelr::add_residuals(model_bad) 

bad_data |> 
  ggplot(aes(x = x, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0)
```

:::

## When homoskedasticity is violated

Implications:

- Estimators $\hat{\beta}_0$ and $\hat{\beta}_1$ are still unbiased
- The standard errors are incorrect

Remedies:

- Consider transformations of $Y$
- Consider Weighted least squares

## Diagnosing normality

- Assess visually with Q-Q plot
- Outliers (point with extreme residual) can result in violation of normality

:::{.pull-left}

**Non-normal residuals**

```{r echo = FALSE, message = FALSE}
bad_data <- data.frame(
  x = seq(0, 10, length.out = 100)
) |> 
  mutate(y_perfect = 2 * x + 5) |> 
  mutate(residuals = rgamma(100, shape = 2, rate = 3)) |> 
  mutate(y = y_perfect + residuals)

model_bad <- lm(y ~ x, data = bad_data)

bad_data <- 
  bad_data |> 
  modelr::add_residuals(model_bad) 

ggplot(bad_data, aes(sample = y)) + 
  stat_qq() + 
  stat_qq_line()
```

:::

:::{.pull-right}

**Normal residuals**


```{r echo = FALSE, message = FALSE}
normal_data <- data.frame(y = rnorm(100))

ggplot(normal_data, aes(sample = y)) + 
  stat_qq() + 
  stat_qq_line()
```

:::



## When normality is violated

Implications:

- With no outliers and a large enough sample Central Limit Theorem corrects for lack of normality
- Can have biased predictions for $Y$

Remedies:

- Consider transformations of $Y$
