---
title: "Math 430: Lecture 5a"
subtitle: "Linear Regression Conditions"
author: "Professor Catalina Medina"
execute:
  echo: true
format: 
  revealjs:
    theme: [sky, ../slide-style.scss]
    slide-number: true
---

```{r}
#| echo: false
library(tidyverse)

elephants <- read_csv(
  "http://csuci-math430.github.io/lectures/data/elephants.csv",
  show_col_types = FALSE
) |> 
  janitor::clean_names()
```


##

We discussed how simple linear regression can be used to solve different goals

**Prediction** 

- We want the fitted line so we can use it on future data

**Inference**

- Learn the relationship between $Y$ and $X$ (confidence interval)
- Test if there is a linear relationship between $Y$ and $X$ (hypothesis test)


## 

Recall our linear regression from last week
```{r}
ggplot(elephants, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Height (cm)", y = "Mass (kg)")
```

. . .

*Was this model appropriate?*

## Why would we care if the model is "appropriate"?

::::{.columns}
:::{.column width="50%"}
```{r}
#| echo: false
ggplot(elephants, aes(x = height, y = mass)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Height (cm)", y = "Mass (kg)")
```

:::

:::{.column width="50%"}

- it may be chance that we got a good fit for our sample, but the model might be a bad fit for the population
- our estimates relied on our conditions, so if those don't hold our estimates are meaningless

:::
::::

## Conditions

$$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i, \epsilon_i \overset{iid}{\sim} N(0, \sigma^2)$$

In order of importance:

1. Linearity: Linear relationship between $X$ and $Y$.
2. Homoskedasticity: Constant variance of $\epsilon_i$.
3. Independence: Independence of $\epsilon_i$.
4. Normality: Normally distributed $\epsilon_i$.

# Checking model conditions

## Visual checks versus statistical tests

People have developed some statistical tests to asses some model conditions.

- We want to avoid unnecessary tests because each test we perform introduces some error.

. . .

Alternatively, we will use visual assessments

- There is some level of subjectivity
- We will use experience and reproducibility to be responsible

## Visual checks

```{r}
#| fig-align: center
#| output-location: slide
model_elephant <- lm(mass ~ height, data = elephants)

library(ggfortify)

autoplot(model_elephant)
```

Recall residuals: $e_i = y_i - \hat{y}_i$

## Linearity

```{r}

```

## Homoskedasticity

## Independence

## Normality

As long as we have a large enough sample (~ 30 observations) we are not too concerned because our estimators will be normal 

